Evaluating the Performance and Energy Efficiency of the COSMO-ART model system

### 2014-04-24 paper_enahpc14/Sec2:COSMO-ART.tex
reduces size of paper to 7 pages

\subsection{Model Description}
\label{subsec:1.1}
The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
regional  to  continental scale  model  coupled  online  to the  COSMO
numerical weather  prediction and climate  model \citep{Baldauf-2011}.
It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
aerosol dynamics  and allows the online calculation  of reactive trace
substances and their interaction  with the atmosphere.  Detailed model
description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
  Knote-2013}.

\subsection{Model Setup}
\label{subsec:1.2}
Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected variance for the duration of the Exa2Green project. To define
a baseline,  it was necessary  to find a run-configuration  capable of
being  recreated in  all subsequent  versions.  Here  we give  a brief
overview of the model setup  for our performance and energy efficiency
evaluation.

@@

The meteorological initial and boundary conditions are obtained by the
the ECMWF  global spectral model IFS  with an update  frequency of 3h.
Boundary data  for gas-phase species are taken  from IFS-MOZART output
at  6h temporal resolution.  The model  setup incorporates  34~2-D and
45~3-D  fields to  be written  out every  hour and  is devoid  of data
assimilation methods.

>>>>

\subsection{Model Description}
\label{subsec:1.1}

The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
regional  to  continental scale  model  coupled  online  to the  COSMO
numerical weather  prediction and climate  model \citep{Baldauf-2011}.
It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
aerosol dynamics  and allows the online calculation  of reactive trace
substances and their interaction  with the atmosphere.  Detailed model
description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
  Knote-2013}.

\subsection{Model Setup}
\label{subsec:1.2}

Establishing effective energy performance benchmarking of a code under
intense development such as COSMO-ART is a challenging task because of
the  absolute necessity that  results must  be reproducible  within an
expected variance for the duration of the Exa2Green project. To define
a baseline,  it was necessary  to find a run-configuration  capable of
being  recreated in  all subsequent  versions.  Here  we give  a brief
overview of the model setup  for our performance and energy efficiency
evaluation.

@@

The meteorological initial and boundary conditions are obtained by the
the ECMWF  global spectral model IFS  with an update  frequency of 3h.
Boundary data  for gas-phase species are taken  from IFS-MOZART output
at  6h temporal resolution.   The model  setup incorporates  34~2D and
45~3D  fields to  be written  out  every hour  and is  devoid of  data
assimilation methods.

### 2014-04-24 paper_enahpc14/Sec4:Experimental-results.tex
reduces size of paper to 7 pages

In  this section,  we  describe a  performance  and energy  efficiency
evaluation  of  different  achitectures  when  running  the  COSMO-ART
baseline.  We  start by  specifying our measurement  methodology along
with the metrics used to  analyse the results on all platforms.  Then,
we detail the environment  setup gathering the considered HPC systems,
the  software  environment and  the  run  configuration.  Finally,  we
discuss benchmark  results and  power-performance traces of  the model
system.

@@

We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
TTS refers to  the total wall clock time  of the application execution
time. ETS  is the amount of  energy spent to  achieve results.  Energy
consumption  is  assessed by  sampling  the  instantaneous peak  power
during execution which  is then averaged and multiplied  by the TTS to
determine  ETS.   Whenever   possible,  multiple  production  runs  of
COSMO-ART  were performed  to  illustrate the  reproducibility of  the
baseline,  and quantify  the  significant uncertainties  in the  power
measurement, as dictated by the available technology.

@@

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
    \caption{Monch: Isola E1 Rack 2 Total Power}
    \label{fig:1}
  \end{center}
\end{figure}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
    \caption{Pilatus: Isola HD Total Power}
    \label{fig:2}
  \end{center}
\end{figure}

Figures~\ref{fig:1} and  \ref{fig:2} account respectively  for Monch's
Isola E1  Rack 2  and Pilatus' Isola  HD total power  measurements for
1-day or 2-days simulations. On  the Intel Ivy Bridge EP based cluster
(i.e. Monch), the 1-day simulation  was issued only twice due to usage
restrictions. As time resolution was set to one update every 5 minutes
for  power sampling,  the average  power consumption  was  computed by
considering 6 values for each single  run.  On the Intel Xeon E5 based
cluster (i.e.   Pilatus), the 1-day  simulation was issued  four times
and a 2-days  run only once. Similarly, the  average power consumption
was computed by  considering 4 values for each single  1-day run and 9
values  for the  2-days  run. Corresponding  results  are gathered  in
Table~\ref{tab:3}.

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.   The reason for  that is  two-fold: (1)  it has  higher clock
frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
at  computing   speed  regardless   to  energy  consumption.   In  our
experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
reducing the energy consumption of Xeon E5 by approximately $7\%$.

@@

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
  \caption{Time-to-solution and energy-to-solution on Tintorrum -
    Polling mode}
  \label{fig:4}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
  \caption{Time-to-solution and energy-to-solution on Tintorrum -
    Polling mode}
  \label{fig:5}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-3.eps}
  \caption{Time-to-solution and Power on Tintorrum - Polling mode}
  \label{fig:6}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-4.eps}
  \caption{Time-to-solution and Power on Tintorrum - Blocking mode}
  \label{fig:7}
\end{figure}

@@

\begin{figure*}[htbf]
  \centering
  \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
  \caption{24 hours simulation trace using the MPI blocking mode on Tintorrum.}
  \label{fig:9}
\end{figure*}

\begin{figure*}[htbf]
  \centering
  \scriptsize
  MPI polling mode\\
  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
  MPI blocking mode\\
  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}
  \caption{1 hour simulation (11h-12h) trace using the MPI blocking mode on Tintorrum.}
  \label{fig:10}
\end{figure*}

\begin{figure*}[htbf]
  \centering
  \scriptsize
  MPI polling mode\\
  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
  MPI blocking mode\\
  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}
  \caption{1 hour simulation (23h-24h) trace using the MPI blocking mode on Tintorrum.}
  \label{fig:11}
\end{figure*}

>>>>

In this section, we present our measurement methodology along with the
metrics considered and discuss energy efficiency and power-performance
traces of the model system.

@@

We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
TTS refers to  the total wall clock time  of the application execution
time. ETS  is the amount of  energy spent to  achieve results.  Energy
consumption  is  assessed by  sampling  the  instantaneous peak  power
during execution which  is then averaged and multiplied  by the TTS to
determine  ETS.   Whenever  possible,  multiple production  runs  were
performed  to  illustrate the  reproducibility  of  the baseline,  and
quantify the  significant uncertainties  in the power  measurement, as
dictated by the available technology.

@@

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
    \caption{\textsc{Monch}: Isola E1 Rack 2 Total Power}
    \label{fig:1}
  \end{center}
\end{figure}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
    \caption{\textsc{Pilatus}: Isola HD Total Power}
    \label{fig:2}
  \end{center}
\end{figure}

Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
\textsc{Monch}'s Isola E1 Rack  2 and \textsc{Pilatus}' Isola HD total
power measurements for  1-day or 2-days simulations. On  the Intel Ivy
Bridge EP  based cluster (i.e.  \textsc{Monch}),  the 1-day simulation
was issued  only twice due  to usage restrictions. As  time resolution
was set to one update every  5 minutes for power sampling, the average
power consumption was computed by considering 6 values for each single
run.  On the Intel Xeon E5 based cluster (i.e.  \textsc{Pilatus}), the
1-day simulation  was issued  four times and  a 2-days run  only once.
Similarly, the average power consumption was computed by considering 4
values  for  each  single  1-day  run  and 9  values  for  the  2-days
run. Corresponding results are gathered in Table~\ref{tab:3}.

@@

In  Figure~\ref{fig:3}, we  compare both  TTS (right  y-axis)  and ETS
(left  y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
outperforms Ivy Bridge EP, being  roughly 1.3x faster.  The reason for
that is  two-fold: (1) it has  higher clock frequency  than Ivy Bridge
(2.6  GHz  against  2.2 GHz),  and  (2)  it  aims at  computing  speed
regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
showed the best energy-to-solution, reducing the energy consumption of
Xeon E5 by approximately $7\%$.

@@

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
  \caption{Time-to-solution and energy-to-solution on
    \textsc{Tintorrum} - Polling mode}
  \label{fig:4}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
  \caption{Time-to-solution and energy-to-solution on
    \textsc{Tintorrum} - Polling mode}
  \label{fig:5}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-3.eps}
  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Polling
    mode}
  \label{fig:6}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-4.eps}
  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Blocking
    mode}
  \label{fig:7}
\end{figure}

@@

\begin{figure*}[htbf]
  \centering
  \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
  \caption{24 hours simulation trace using the MPI blocking mode on
    \textsc{Tintorrum}.}
  \label{fig:9}
\end{figure*}

\begin{figure*}[htbf]
  \centering
  \scriptsize
  MPI polling mode\\
  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
  MPI blocking mode\\
  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}
  \caption{1 hour simulation (11h-12h) trace using the MPI blocking
    mode on \textsc{Tintorrum}.}
  \label{fig:10}
\end{figure*}

\begin{figure*}[htbf]
  \centering
  \scriptsize
  MPI polling mode\\
  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
  MPI blocking mode\\
  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}
  \caption{1 hour simulation (23h-24h) trace using the MPI blocking
    mode on \textsc{Tintorrum}.}
  \label{fig:11}
\end{figure*}
